---
title: "gwas_assessment"
author: "Silpa"
date: "2025-10-28"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(data.table)
library(readr)
library(ggplot2)
library(reshape2)
library(patchwork)
library(tidyr)
library(qqman)
```

## Genotype dataset and Phenotype distribution

```{r}
# read bim file 
alz_phenotype <- fread("~/Documents/GitHub/5BD002/gwas_omics_assessment/ALZ/ALZ_sample.fam",
             select = c(2,6), header = FALSE, col.names = c("IID", "PhenoTypeValue"))

alz_variants <- fread("~/Documents/GitHub/5BD002/gwas_omics_assessment/ALZ/ALZ_sample.bim",
             select = c(1,2,4,5,6), header = FALSE, col.names = c("ChrCode", "VarID", "BasePairCoord", "Allele1", "Allele2"))
```

```{r}
ggplot(alz_phenotype, aes(factor(PhenoTypeValue))) +
  geom_bar(fill = "steelblue") +
  geom_text(stat = "count", aes(label = ..count..), vjust = -0.3) +
  labs(
    x = "Phenotype (1 = Control, 2 = Case)",
    y = "Count",
    title = "Phenotype Distribution"
  ) +
  theme_minimal(base_size = 12)
```

## PCA 

run bash : 

./plink2 --silent  --bfile ALZ_sample --maf 0.05  --indep-pairwise 500kb 0.1  --out ALZ_indep

./plink2 --silent  --bfile ALZ_sample  --extract ALZ_indep.prune.in  --pca --out ALZ_pca

```{r}
alz_pca <- fread("~/Documents/GitHub/5BD002/gwas_omics_assessment/ALZ/ALZ_pca.eigenvec")
alz_merged <- merge(alz_pca, alz_phenotype[, c("IID", "PhenoTypeValue")],
                    by = c("IID"))

ggplot(alz_merged, aes(PC1, PC2, color = factor(PhenoTypeValue))) +
  geom_point(alpha = 0.5, size = 1.8) +
  labs(color = "Phenotype (1 = Control, 2 = Case)",
       title = "PCA of Genotype Data (colored by Case/Control)")

# does the phenotype vary by "ancestry" (PCs)
ggplot(alz_merged, aes(PC1, PhenoTypeValue)) + geom_point() + geom_smooth()
```

Show all 10 PCA plots with phenotype
```{r}
m_long <- alz_merged %>%
  pivot_longer(
    cols = starts_with("PC"),
    names_to = "PC",
    values_to = "PC_value"
  )
ggplot(m_long, aes(PC_value, PhenoTypeValue, color = factor(PhenoTypeValue))) +
  geom_point(alpha = 0.7, size = 1) +
  geom_smooth(se = FALSE, color = "blue") +
  facet_wrap(~ PC, ncol = 5) +     # <---- 2 rows × 5 columns
  labs(
    x = "Principal Component Value",
    y = "Phenotype (1 = Control, 2 = Case)",
    title = "Relationship between PCA Components and Phenotype"
  )
```

Show the statistical association of PCAs with Phenotype:

```{r}
alz_merged$Pheno01 <- ifelse(alz_merged$PhenoTypeValue == 2, 1, 0)
PCs_fit_Phenotype <- glm(Pheno01 ~ PC1 + PC2 + PC3 + PC4 + PC5 + PC6 + PC7 + PC8 + PC9 + PC10, data = alz_merged, family = binomial())
summary(PCs_fit_Phenotype)
```


## GWAS without PCA adjustment

run : ./plink2 --silent --bfile ALZ_sample --glm allow-no-covars hide-covar  --out ALZ_GWAS_noPC

```{r}
gwas_unadj <- fread("~/Documents/GitHub/5BD002/gwas_omics_assessment/ALZ/ALZ_GWAS_noPC.PHENO1.glm.logistic.hybrid")
gwas_unadj[, P := pmax(1e-20, P)]
manhattan(gwas_unadj, chr = "#CHROM", bp = "POS", snp = "ID", p = "P",
          main = "GWAS Manhattan Plot (No PC Adjustment)")
```

```{r}
qq(gwas_unadj$P)
```


## GWAS with PCA adjustment

```{r}
```

## LD clumping of GWAS results

```{r}
```

